---
title: "R Notebook"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

```{r load, include=FALSE}
required_pkgs <- c(
  "DBI",
  "RSQLite",
  "biomaRt",
  "dplyr",
  "edgeR",
  "gprofiler2",
  "GSA",
  "BiocManager",
  "RCy3",
  "httr",
  "tidyr",
  "stringr"
)

for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}
```

This is part of the same process as A1.4 for we stored the data before
hypothesis testing.

```{r reload-datasets}
## ---- load-a1-objects ----

obj <- readRDS("data/d1_qc_objects.rds")

dge  <- obj$dge_filtered
meta <- obj$meta
mappings <- obj$mappings
```

Recreate design matrix

```{r recreate design}
## ---- recreate-design ----

meta$timepoint <- factor(meta$timepoint)
meta$treatment <- factor(meta$treatment, levels = c("CTL", "PAL"))

design <- model.matrix(~ timepoint + treatment, data = meta)
```

Re-estimate dispersion and fit QL model

```{r refit QL}
## ---- fit-edger-model ----

dge <- estimateDisp(dge, design)

fit <- glmQLFit(dge, design)
```

Verify if its working as before:

```{r BCV}
plotBCV(dge)
```

Hypoothesis testing

```{r ql-test}
## ---- ql-test ----

qlf_pal_vs_ctl <- glmQLFTest(
  fit,
  coef = "treatmentPAL"
)

res_pal <- topTags(qlf_pal_vs_ctl, n = Inf)$table %>%
  tibble::rownames_to_column("ensembl")
```

Verify same number as before:

```{r }
sum(res_pal$PValue < 0.05)
```

```{r}
sig_res <- res_pal %>%
  filter(FDR < 0.05, abs(logFC) >= 1)

nrow(sig_res)
```

Split into up and down:

```{r}
sig_up <- sig_res %>% filter(logFC > 0)
sig_down <- sig_res %>% filter(logFC < 0)

nrow(sig_up)
nrow(sig_down)
```

1.1 Method Choice Method: Thresholded Over-Representation Analysis (ORA)

We use:

g:Profiler (gprofiler2 R package)

Hypergeometric test

Benjamini–Hochberg FDR correction

Why this method?

We already defined significant genes using FDR \< 0.05 and \|logFC\| ≥
1.

ORA is appropriate when using a pre-thresholded gene list.

It tests whether gene sets are enriched beyond random expectation.

We use g:Profiler built-in annotation sources:

```{r Sources}
gprofsources = c("GO:BP", "REAC", "WP")
```

Enrichment analysis was performed using g:Profiler (version 2025-03-16)
with Homo sapiens annotation databases including GO (BP, MF, CC),
Reactome, WikiPathways, KEGG, transcription factor targets, microRNA
targets, CORUM complexes, Human Protein Atlas, and Human Phenotype
Ontology.

Annotation version

```{r Report Version, include = FALSE}
get_version_info(organism = "hsapiens")
```

Inspect gene lists

```{r inspect gene lists, include = FALSE}
head(sig_up)
```

Extract the Gene list by unique ensembl or biomart symbel.

```{r Create gene symbel only list}
genes_up   <- unique(sig_up$ensembl)
genes_down <- unique(sig_down$ensembl)
genes_all  <- unique(sig_res$ensembl)

length(genes_up)
length(genes_down)
```

Run Upregulated

```{r up}
gost_up <- gost(
  query = genes_up,
  organism = "hsapiens",
  significant = FALSE,
  ordered_query = FALSE,
  exclude_iea = TRUE,
  correction_method = "fdr",
  sources = gprofsources
)

res_up <- gost_up$result
```

Run Downregulated

```{r down}
gost_down <- gost(
  query = genes_down,
  organism = "hsapiens",
  significant = FALSE,
  ordered_query = FALSE,
  exclude_iea = TRUE,
  correction_method = "fdr",
  sources = gprofsources
)

res_down <- gost_down$result
```

Combined

```{r all}
gost_all <- gost(
  query = genes_all,
  organism = "hsapiens",
  significant = FALSE,
  ordered_query = FALSE,
  exclude_iea = TRUE,
  correction_method = "fdr",
  sources = gprofsources
)

res_all <- gost_all$result
```

We performed thresholded over-representation analysis (ORA) using the
gprofiler2::gost() function with the following parameters:

query = genes_up / genes_down / genes_all

The query represents the list of significantly differentially expressed
genes identified from edgeR analysis (FDR \< 0.05 and \|log2FC\| ≥ 1).
Up-regulated, down-regulated, and combined gene sets were analyzed
separately to preserve biological directionality.

organism = "hsapiens"

The organism parameter specifies that Homo sapiens annotation databases
were used for enrichment analysis.

significant = FALSE

We set significant = FALSE to retrieve all enrichment results rather
than restricting output to only those passing g:Profiler’s internal
significance threshold. This allows us to apply our own filtering
criteria downstream.

ordered_query = FALSE

Because this analysis is thresholded ORA (not ranked GSEA), genes were
treated as an unranked set. Therefore, ordered_query = FALSE was
appropriate.

exclude_iea = TRUE

Inferred Electronic Annotations (IEA) were excluded to increase
annotation reliability by limiting enrichment to manually curated or
experimentally supported terms.

correction_method = "fdr"

Multiple hypothesis correction was performed using the
Benjamini–Hochberg False Discovery Rate (FDR) method to control for
false positives across many tested gene sets.

sources = gprofsources

The following annotation databases were included:

<GO:BP> (Biological Process)

<GO:MF> (Molecular Function)

<GO:CC> (Cellular Component)

REAC (Reactome pathways)

WP (WikiPathways)

KEGG

TF (Transcription factor targets)

MIRNA (miRNA target sets)

CORUM (protein complexes)

HPA (Human Protein Atlas)

HP (Human Phenotype Ontology)

Including multiple annotation sources allows interrogation of biological
processes, pathways, regulatory networks, and disease associations.

Analysis:

The orginal paper applies RAIN algorithm for seperating cirardian genes
from other effected genes as lipid intake alters well studied pathway
such as lipid intake. Here we see if a bulk enrichment non theresholded
ora can identify cirardian genes alone with lipid intake.

1.  Identify the biological expectation from the paper

From the paragraph:

“Palmitate-responsive genes were associated with gene ontology pathways
related to lipid metabolism (Figure 1E).”

So the expected pathways include things like:

Typical lipid-related GO BP terms:

lipid metabolic process

fatty acid metabolic process

fatty acid beta oxidation

lipid biosynthetic process

sterol metabolic process

lipid transport

These become your benchmark.

If ORA finds these → result is relevant.

```{r keyword rules}
classify_theme <- function(df) {
  df %>%
    mutate(
      theme = case_when(
        str_detect(term_name, regex("circadian|clock|rhythm", ignore_case = TRUE)) ~ "Circadian",
        str_detect(term_name, regex("lipid|fatty|cholesterol|sterol|acyl", ignore_case = TRUE)) ~ "Lipid metabolism",
        TRUE ~ "Other"
      )
    )
}
```

```{r Apply classification}
res_up_theme   <- classify_theme(res_up)
res_all_theme  <- classify_theme(res_all)
```

```{r count}
count_theme <- function(df, label) {
  df %>%
    group_by(theme) %>%
    summarise(count = n(), .groups = "drop") %>%
    mutate(group = label)
}

summary_up   <- count_theme(res_up_theme, "Upregulated")
summary_all  <- count_theme(res_all_theme, "All")
```

```{r threshold summery}
threshold_summary <- bind_rows(
  summary_up,
  summary_all
)

themes <- c("Lipid metabolism", "Circadian", "Other")
groups <- c("Upregulated", "Downregulated", "All")

threshold_summary <- threshold_summary %>%
  complete(group = groups, theme = themes, fill = list(count = 0))
```

```{r plot}
library(ggplot2)

ggplot(threshold_summary, aes(x = group, y = count, fill = theme)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(
    title = "Enriched genesets by biological theme (Thresholded ORA)",
    x = "Gene set",
    y = "Number of genesets",
    fill = "Theme"
  )
```

figure.x Summery of gene set counts for up regulated genesets, down
regulated genesets and both gene sets combined.

```{r table of top term}
get_top_terms <- function(df, group_name) {
  df %>%
    group_by(theme) %>%
    slice_min(p_value, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    select(theme, top_term = term_name, p_value) %>%
    mutate(group = group_name)
}

top_up  <- get_top_terms(res_up_theme, "Upregulated")
top_all <- get_top_terms(res_all_theme, "All")

top_terms <- bind_rows(top_up, top_all)

themes <- c("Lipid metabolism", "Circadian", "Other")
groups <- c("Upregulated", "Downregulated", "All")

top_terms <- top_terms %>%
  complete(group = groups, theme = themes,
           fill = list(top_term = NA, p_value = NA))

threshold_summary_table <- threshold_summary %>%
  left_join(top_terms, by = c("group", "theme"))

knitr::kable(
  threshold_summary_table,
  caption = "Summary of enriched pathways by biological theme with top representative term"
)
```

Table x. Summery of top term and significant level for each theme.

Interpertation: the g:profiling by fdr cutoff o.o5 did not identify any
significant circardian gene set, but identified 53 lipid metabolism
related geneset and 579 other geneset in using the positive experssed
gene list. Using both positive and negative gene together did not effect
the results as no significant gene set for 3 negative regulated gene
added. see table x, figure x

Size filtered Results reusing similar code

```{r filter size}
min_gs_size <- 3
max_gs_size <- 250
min_intersection <- 2

res_up_f <- subset(res_up,
                   term_size >= min_gs_size &
                   term_size <= max_gs_size &
                   intersection_size >= min_intersection)

res_all_f <- subset(res_all,
                    term_size >= min_gs_size &
                    term_size <= max_gs_size &
                    intersection_size >= min_intersection)

classify_theme <- function(df) {
  df %>%
    mutate(
      theme = case_when(
        str_detect(term_name, regex("circadian|clock|rhythm", ignore_case = TRUE)) ~ "Circadian",
        str_detect(term_name, regex("lipid|fatty|cholesterol|sterol|acyl", ignore_case = TRUE)) ~ "Lipid metabolism",
        TRUE ~ "Other"
      )
    )
}

res_up_f_theme  <- classify_theme(res_up_f)
res_all_f_theme <- classify_theme(res_all_f)

count_theme <- function(df, label) {
  df %>%
    group_by(theme) %>%
    summarise(count = n(), .groups = "drop") %>%
    mutate(group = label)
}

summary_up_f  <- count_theme(res_up_f_theme, "Upregulated")
summary_all_f <- count_theme(res_all_f_theme, "All")

filtered_summary <- bind_rows(summary_up_f, summary_all_f)

get_top_terms <- function(df, group_name) {
  df %>%
    group_by(theme) %>%
    slice_min(p_value, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    select(theme, top_term = term_name, p_value) %>%
    mutate(group = group_name)
}

top_up_f  <- get_top_terms(res_up_f_theme, "Upregulated")
top_all_f <- get_top_terms(res_all_f_theme, "All")

top_terms_f <- bind_rows(top_up_f, top_all_f)

themes <- c("Lipid metabolism", "Circadian", "Other")
groups <- c("Upregulated", "All")

filtered_summary <- filtered_summary %>%
  complete(group = groups, theme = themes, fill = list(count = 0))

top_terms_f <- top_terms_f %>%
  complete(group = groups, theme = themes,
           fill = list(top_term = NA, p_value = NA))

filtered_summary_table <- filtered_summary %>%
  left_join(top_terms_f, by = c("group", "theme"))

knitr::kable(
  filtered_summary_table,
  caption = "Filtered ORA pathway summary with representative term per biological theme"
)
```

Table Y. Top term anhd count for each theme after filtering geneset size
3\<size\<200.

The result shows less lipid metabolism for less general pathways after
filtering with a smaller total gene. Ther are 8 for lipid metabolism, 52
for other and still 0 for circardian genes.

In the paper it was conclued as large fraction of cycling genes stops
cycling after palmitate treatment, whle only influenced the number of
rhythmic genes, with minimal effect on expression of clock machenery
components. By stopping cyclic behavior but maintain a similar
experssion level, it aligns with our result of the insignificant of the
cyclic genes' diffrenetial expression.

Part 2 Unthrreshold set output location:

```{r output location}
working_dir <- "./data"
```

create ranked gene list

```{r ranked gene list}
library(dplyr)

# Create ranked table
res_rank <- res_pal %>%
  mutate(rank_score = sign(logFC) * -log10(PValue)) %>%
  arrange(desc(rank_score))

# Keep only required columns
rnk_df <- res_rank %>%
  select(ensembl, rank_score)

# Define file path
working_dir <- "./data"
dir.create(working_dir, showWarnings = FALSE)

rnk_file <- "PAL_vs_CTL_ranked.rnk"
rnk_path <- file.path(working_dir, rnk_file)

# Write .rnk file
write.table(
  rnk_df,
  file = rnk_path,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE,
  quote = FALSE
)
```

Download fixed version fir GMT from baderlab

```{r download GMT}
library(RCurl)

output_dir <- "./generated_data/gsea"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

gmt_url  <- "https://download.baderlab.org/EM_Genesets/September_01_2025/Human/symbol/Human_GOBP_AllPathways_noPFOCR_no_GO_iea_September_01_2025_symbol.gmt"

gmt_file <- "Human_GOBP_AllPathways_noPFOCR_no_GO_iea_September_01_2025_symbol.gmt"
gmt_path <- file.path(output_dir, gmt_file)

if (!file.exists(gmt_path)) {
  download.file(gmt_url, destfile = gmt_path, mode = "wb")
}
```

September 01, 2025 GO BP + Pathways No IEA Human symbols

Configure GSEA CLI

```{r }
gsea_jar <- path.expand("~/GSEA_4.4.0/gsea-cli.sh")
```

Run:

```{r run GSEA}
analysis_name <- "PAL_vs_CTL"

gsea_stdout <- file.path(output_dir, "gsea_output.txt")

command <- paste(
  shQuote(gsea_jar),
  "GSEAPreRanked",
  "-gmx", shQuote(gmt_path),
  "-rnk", shQuote(rnk_path),
  "-collapse false",
  "-nperm 1000",
  "-scoring_scheme weighted",
  "-rpt_label", shQuote(analysis_name),
  "-plot_top_x 20",
  "-rnd_seed 12345",
  "-set_max 200",
  "-set_min 15",
  "-zip_report false",
  "-out", shQuote(output_dir),
  ">", shQuote(gsea_stdout)
)

cat("Running command:\n", command, "\n")
system(command)
```

Gene sets were restricted to sizes between 15 and 200 genes to exclude
very small unstable sets and overly broad non-specific categories.

Summary of enrichment results

The GSEA analysis identified several strongly enriched pathways
associated with lipid metabolism and metabolic stress responses.

Among the positively enriched pathways were processes related to:

mTORC1 signaling

cholesterol biosynthesis regulation

lipid biosynthetic processes

PPAR-α regulated lipid metabolism

steroid metabolism

endoplasmic reticulum stress response

These pathways showed high normalized enrichment scores (NES ≈ 2.3–2.5)
and strong statistical significance.

gsea_report_for_na_pos_17731128…

Several metabolic pathways related to mitochondrial respiration and
oxidative phosphorylation were negatively enriched.

gsea_report_for_na_neg_17731128…

Overall, the enrichment pattern indicates that palmitate treatment
strongly affects lipid metabolic pathways and cellular stress responses,
which is consistent with the known lipotoxic effects of saturated fatty
acids.

Circadian-related pathways were detected only weakly and were not among
the most significant results. For example, circadian clock related gene
sets appeared with relatively low enrichment scores and non-significant
FDR values.

The non-thresholded GSEA analysis provides a complementary perspective
because it considers all genes ranked by differential expression rather
than only a subset passing significance thresholds.

Compared with the thresholded ORA:

GSEA identified stronger enrichment of coordinated metabolic pathways

Lipid metabolism and cholesterol biosynthesis pathways were clearly
detected

Additional metabolic stress pathways such as ER stress and mTOR
signaling were also revealed

However, circadian pathways remained weakly enriched even in the
non-thresholded analysis.

Therefore, while both approaches highlight metabolic responses to
palmitate, GSEA captures broader coordinated pathway signals that may be
missed by thresholded ORA.

Is this a straightforward comparison?

The comparison between thresholded ORA and non-thresholded GSEA is not
entirely straightforward, because the two methods rely on fundamentally
different assumptions.

ORA tests whether a selected list of significant genes is
over-represented in known pathways, which depends heavily on the chosen
significance threshold.

In contrast, GSEA evaluates whether genes belonging to a pathway tend to
appear toward the top or bottom of the ranked gene list, allowing
detection of subtle but coordinated expression changes across many
genes.

As a result, GSEA can identify biologically meaningful pathways even
when individual genes do not pass strict differential expression
thresholds.

Thus, while both analyses point toward lipid metabolism as the dominant
biological response to palmitate, GSEA provides a more comprehensive
view of coordinated pathway regulation.
