---
title: "R Notebook"
output: html_notebook
---
```{r load, include=FALSE}
required_pkgs <- c(
  "DBI",
  "RSQLite",
  "biomaRt",
  "dplyr",
  "edgeR",
  "gprofiler2",
  "GSA"
)

for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}
```
This is part of the same process as A1.4 for we stored the data before hypothesis testing.
```{r reload-datasets}
## ---- load-a1-objects ----

obj <- readRDS("data/d1_qc_objects.rds")

dge  <- obj$dge_filtered
meta <- obj$meta
mappings <- obj$mappings
```
Recreate design matrix
```{r recreate design}
## ---- recreate-design ----

meta$timepoint <- factor(meta$timepoint)
meta$treatment <- factor(meta$treatment, levels = c("CTL", "PAL"))

design <- model.matrix(~ timepoint + treatment, data = meta)

colnames(design)
```
Re-estimate dispersion and fit QL model
```{r refit QL}
## ---- fit-edger-model ----

dge <- estimateDisp(dge, design)

fit <- glmQLFit(dge, design)
```
Verify if its working as before:
``` {r BCV}
plotBCV(dge)
```
Hypoothesis testing
```{r ql-test}
## ---- ql-test ----

qlf_pal_vs_ctl <- glmQLFTest(
  fit,
  coef = "treatmentPAL"
)

res_pal <- topTags(qlf_pal_vs_ctl, n = Inf)$table %>%
  tibble::rownames_to_column("ensembl")
```
Verify same number as before:
```{r }
sum(res_pal$PValue < 0.05)
```
```{r}
sig_res <- res_pal %>%
  filter(FDR < 0.05, abs(logFC) >= 1)

nrow(sig_res)
```
Split into up and down:
```{r}
sig_up <- sig_res %>% filter(logFC > 0)
sig_down <- sig_res %>% filter(logFC < 0)

nrow(sig_up)
nrow(sig_down)
```

1.1 Method Choice
Method: Thresholded Over-Representation Analysis (ORA)

We use:

g:Profiler (gprofiler2 R package)

Hypergeometric test

Benjamini–Hochberg FDR correction

Why this method?

We already defined significant genes using FDR < 0.05 and |logFC| ≥ 1.

ORA is appropriate when using a pre-thresholded gene list.

It tests whether gene sets are enriched beyond random expectation.

This is NOT GSEA (ranked analysis). This is classic ORA.

We use g:Profiler built-in annotation sources:
``` {r Sources}
gprofsources = c("GO:BP", "REAC", "WP")
```

Enrichment analysis was performed using g:Profiler (version X.X) with Homo sapiens annotation databases including GO (BP, MF, CC), Reactome, WikiPathways, KEGG, transcription factor targets, microRNA targets, CORUM complexes, Human Protein Atlas, and Human Phenotype Ontology.

Annotation version
```{r Report Version}
get_version_info(organism = "hsapiens")
```
```{r inspect gene lists}
head(sig_up)
```
```{r Create gene symbel only list}
genes_up   <- unique(sig_up$ensembl)
genes_down <- unique(sig_down$ensembl)
genes_all  <- unique(sig_res$ensembl)

length(genes_up)
length(genes_down)
```
Upregulated
```{r up}
gost_up <- gost(
  query = genes_up,
  organism = "hsapiens",
  significant = FALSE,
  ordered_query = FALSE,
  exclude_iea = TRUE,
  correction_method = "fdr",
  sources = gprofsources
)

res_up <- gost_up$result
```
Down
```{r down}
gost_down <- gost(
  query = genes_down,
  organism = "hsapiens",
  significant = FALSE,
  ordered_query = FALSE,
  exclude_iea = TRUE,
  correction_method = "fdr",
  sources = gprofsources
)

res_down <- gost_down$result
```
Combined
```{r all}
gost_all <- gost(
  query = genes_all,
  organism = "hsapiens",
  significant = FALSE,
  ordered_query = FALSE,
  exclude_iea = TRUE,
  correction_method = "fdr",
  sources = gprofsources
)

res_all <- gost_all$result
```
We performed thresholded over-representation analysis (ORA) using the gprofiler2::gost() function with the following parameters:

🔹 query = genes_up / genes_down / genes_all

The query represents the list of significantly differentially expressed genes identified from edgeR analysis (FDR < 0.05 and |log2FC| ≥ 1).
Up-regulated, down-regulated, and combined gene sets were analyzed separately to preserve biological directionality.

🔹 organism = "hsapiens"

The organism parameter specifies that Homo sapiens annotation databases were used for enrichment analysis.

significant = FALSE

We set significant = FALSE to retrieve all enrichment results rather than restricting output to only those passing g:Profiler’s internal significance threshold.
This allows us to apply our own filtering criteria downstream.

ordered_query = FALSE

Because this analysis is thresholded ORA (not ranked GSEA), genes were treated as an unranked set.
Therefore, ordered_query = FALSE was appropriate.

exclude_iea = TRUE

Inferred Electronic Annotations (IEA) were excluded to increase annotation reliability by limiting enrichment to manually curated or experimentally supported terms.

correction_method = "fdr"

Multiple hypothesis correction was performed using the Benjamini–Hochberg False Discovery Rate (FDR) method to control for false positives across many tested gene sets.

sources = gprofsources

The following annotation databases were included:

GO:BP (Biological Process)

GO:MF (Molecular Function)

GO:CC (Cellular Component)

REAC (Reactome pathways)

WP (WikiPathways)

KEGG

TF (Transcription factor targets)

MIRNA (miRNA target sets)

CORUM (protein complexes)

HPA (Human Protein Atlas)

HP (Human Phenotype Ontology)

Including multiple annotation sources allows interrogation of biological processes, pathways, regulatory networks, and disease associations.

Size filtered Results
``` {r filter size}
min_gs_size <- 3
max_gs_size <- 250
min_intersection <- 2

res_up_f <- subset(res_up,
                   term_size >= min_gs_size &
                   term_size <= max_gs_size &
                   intersection_size >= min_intersection)

res_down_f <- subset(res_down,
                     term_size >= min_gs_size &
                     term_size <= max_gs_size &
                     intersection_size >= min_intersection)

res_all_f <- subset(res_all,
                    term_size >= min_gs_size &
                    term_size <= max_gs_size &
                    intersection_size >= min_intersection)
```
